---
title: "In-class-template"
author: "Macey Hartmann"
format: pdf
---

# 1. Set Up 

## Packages 

```{r}
#| label: packages

# general use and cleaning 
library(tidyverse)
library(here)
library(janitor)
library(snakecase)

# wrap axis labels 
library(scales)

# reading .xlsx files 
library(readxl)

# visualize missing data 
library(naniar)

# arrange plots 
library(patchwork)

```

## Data

```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"), 
  sheet = "Aquatic Insects"
)

# site information
sites <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "sites"
)

# water quality 
water_quality <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Water Quality", 
  na = c("", "n/a", "N/A", "over", "173+")
)

```

# 2. Cleaning

## 'sites' object

Goal: create a new object from "sites" that only inlcudes NCOS sites

```{r}

# creating new object from sites
ncos_sites <- sites |> 
  # filtering to only includes NCOS quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")

```



## clean 'taxon_list'

Goal: convert character strings in the existing 'taxon_list' data frame into snakecase

```{r}

# creating new object from taxon_list 
taxon_list_clean <- taxon_list |>
  # converted taxon names to snake case 
  mutate(verbatim_name = to_any_case(verbatim_name, 
                                     case = "snake"))
```

## clean 'water_quality'

Goal: create a data frame with mean pH, mean DO (mg/L), mean salinity (ppt) for each date and site 

- create a new object from 'water_quality' 
- clean column names 
- only include surveys from quarterly NCOS sites 
- create a new column called date_site to join with later data frames 
- group by date_site 
- calculate median pH, salinity, and dissolved oxygen 
- ungroup data frame 
- filter out any outliers 

```{r}

# creating new object from water quality
water_quality_clean <- water_quality |> 
  # cleaning column names
  clean_names() |> 
  # filtering join: only include sites in ncos_sites 
  semi_join(ncos_sites, by = "site") |> 
  # creating new 
  unite("date_site", date, site, remove = TRUE) |> 
  group_by(date_site) |> 
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE), 
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  ungroup() |> 
  # filtering out salinity outliers 
  filter(date_site != "2022-08-12_NVBR")


```

## clean 'aq_ins'

Goal: create a long format data frame in which each row contains the counted individuals in each taxon during each survey. The data frame should include taxonomic information for each organism and the water quality information for each survey

- create a new object from 'aq_ins'
- clean column names 
- filter to only include quarterly NCOS sites 
- select columns 
- filter to only include "filtered beaker" sample 
- convert the data frame to long format 
- group by site, date, and taxon 
- calculate average abundance per liter (sum counts divided by 7.5 liters)
- create a new column called 'date_site' to join with water quality
- join data frames 
- join with taxon list 

```{r}

aq_ins_clean <- aq_ins |> 
  clean_names() |> 
  semi_join(ncos_sites, by = "site") |> 
  rename(date = date_on_vial) |> 
  select(date, sample_type, site, 
         ostracod, 
         copepod,
         hemiptera_corixidae_boatman, 
         cladocera, 
         nematode, 
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid, 
         annelida_polychaete) |> 
  filter(sample_type %in% c("FB250", "FB 250")) |> 
  pivot_longer( 
    # columns to make longer
    cols = ostracod:annelida_polychaete, 
    # new column for the old column names
    names_to = "taxon",
    # new column for the values in each of the old columns 
    values_to = "abundance"
    ) |> 
  group_by(date, site, taxon) |> 
  summarize(ave_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  ungroup() |> 
  unite("date_site", date, site, remove = FALSE) |> 
  left_join(water_quality_clean, by = "date_site") |> 
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "South of Venoco Bridge",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek" 
  ))
  
```






